### 2016-06-15 paper/icfp16-cameraready/evaluation.tex
finish draft of user study

\paragraph{Evaluation Goals}
%
There are two questions we seek to answer with our evaluation:
%
\begin{enumerate}
\item \emphbf{Witness Coverage}
      How many ill-typed programs can we find witnesses for?
\item \emphbf{Witness Complexity}
      How \emph{complex} are the traces produced by the witnesses?
\end{enumerate}

\paragraph{Benchmarks}
We answer these questions on two sets of ill-typed programs, \ie\
programs that were rejected by the \ocaml\ compiler because of a
type error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at UC San Diego.
%
We recorded each interaction with the \ocaml\ top-level system over the
course of the first three assignments (IRB
% \# hidden for blind review),
\#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs.
%
The second dataset --- widely used in the literature --- comes from a
graduate-level course at the University of Washington~\cite{Lerner2006-pj},
from which we extracted 284 ill-typed programs.
%
Both datasets contain relatively small programs, the largest being 348
SLoC; however, they demonstrate a variety of functional programming
idioms including (tail) recursive functions, higher-order functions,
polymorphic data types, and expression evaluators.

@@

\begin{figure*}[t]
\centering
\begin{minipage}{\linewidth}
\centering
\includegraphics[width=0.49\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.49\linewidth]{distrib.png}
\end{minipage}
% \vspace{-8ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 79--85\% of
  the programs in under one second, improving slightly to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3--4\% of the time, and
  diverging loops 2--3\% of the time. For the remaining 11--12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure*}

@@

\subsection{Trace Complexity}
\label{sec:trace-complexity}

@@

% \item \ES{others?}
%
\begin{figure*}[t]
\centering
\includegraphics[width=0.49\linewidth]{trace_size_step.png}
\includegraphics[width=0.49\linewidth]{trace_size_jump.png}
\caption{Complexity of the generated traces. 81\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure*}
%

@@

\subsection{User Study}
\label{sec:user-study}
Finally, to test the explanatory power of our jump-compressed traces, we
ran a user study (IRB \#XXXX) at the University of Virginia (UVA).
%
We added four questions to an exam in the Spring session of UVA's
undergraduate Programming Languages course (CS XXX), in which presented
the students with ill-typed \ocaml\ programs and asked them to
%
(1) \emph{explain} the type error, and
%
(2) \emph{fix} the type error.
%
For each problem the student would be given the ill-typed program and
either \ocaml's error message or \toolname's jump-compressed trace.
%
The four problems assigned to the students were the
%
@sumList@, @digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces}
as well as the following @append@ program....\ES{HERE}

We then instructed four annotators (one of whom is an author, the other
three were teaching assistants at UCSD) to classify the answers as
correct or incorrect.

We performed an inter-rater reliability (IRR) analysis to determine the
degree to which the annotators consistently graded the exams.
%
As we had more than two annotators assigning nominal ratings we used
Fleiss' kappa~\cite{XXX} to measure IRR.
%
The resulting kappa values, $\kappa = 0.72$ for the explanations and
$\kappa = 0.83$ for the fixes, indicate substantial agreement among
the annotators~\cite{XXX}.

The results of one of the annotators are shown in
Figure~\ref{fig:results-user-study}.
%
\begin{figure*}[t]
\centering
\includegraphics[width=0.49\linewidth]{user-study-reason.png}
\includegraphics[width=0.49\linewidth]{user-study-fix.png}
\caption{Results of the user study. The students given \toolname's
  jump-compressed trace consistently scored better than those
  given \ocaml's type error.}
\label{fig:results-user-study}
\end{figure*}
%
The students given \toolname's jump-compressed trace were consistently
more likely to correctly explain and fix the type error than those
given \ocaml's error message.

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors: (1) with high coverage in a timespan amenable to
compile-time analysis, and (2) with traces that have a low average
complexity of 7 jumps.

>>>>

\paragraph{Evaluation Goals}
%
There are three questions we seek to answer with our evaluation:
%
\begin{enumerate}
\item \emphbf{Witness Coverage}
      How many ill-typed programs can we find witnesses for?
\item \emphbf{Witness Complexity}
      How \emph{complex} are the traces produced by the witnesses?
\item \emphbf{Witness Utility}
      How \emph{helpful} are the witnesses and traces in debugging type errors?
\end{enumerate}

\paragraph{Benchmarks}
We answer the first two questions on two sets of ill-typed programs,
\ie\ programs that were rejected by the \ocaml\ compiler because of a
type error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at UC San Diego.
%
We recorded each interaction with the \ocaml\ top-level system over the
course of the first three assignments (IRB
% \# hidden for blind review),
\#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs.
%
The second dataset --- widely used in the literature --- comes from a
graduate-level course at the University of Washington~\cite{Lerner2006-pj},
from which we extracted 284 ill-typed programs.
%
Both datasets contain relatively small programs, the largest being 348
SLoC; however, they demonstrate a variety of functional programming
idioms including (tail) recursive functions, higher-order functions,
polymorphic data types, and expression evaluators.

@@

\begin{figure*}[ht]
\centering
\begin{minipage}{\linewidth}
\centering
\includegraphics[width=0.49\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.49\linewidth]{distrib.png}
\end{minipage}
% \vspace{-8ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 79--85\% of
  the programs in under one second, improving slightly to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3--4\% of the time, and
  diverging loops 2--3\% of the time. For the remaining 11--12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure*}

@@

\subsection{Witness Complexity}
\label{sec:trace-complexity}

@@

% \item \ES{others?}
%
\begin{figure*}[ht]
\centering
\includegraphics[width=0.49\linewidth]{trace_size_step.png}
\includegraphics[width=0.49\linewidth]{trace_size_jump.png}
\caption{Complexity of the generated traces. 81\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure*}
%

@@

\subsection{Measuring Witness Utility}
\label{sec:user-study}
Finally, to test the explanatory power of our jump-compressed traces, we
ran a user study (IRB \#XXXX) at the University of Virginia (UVA).
%
We included four problems in an exam in the Spring session of UVA's
undergraduate Programming Languages course (CS XXX).
%
The problems presented the students with ill-typed \ocaml\ programs and
asked them to
%
(1) \emph{explain} the type error, and
%
(2) \emph{fix} the type error.
%
For each problem the student would be given the ill-typed program and
either \ocaml's error message or \toolname's jump-compressed trace.
%
The four problems assigned to the students were the @sumList@,
@digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces},
as well as the following @append@ program
%
\begin{ecode}
  let append x l =
    match x with
    | []   -> l
    | h::t -> h :: t :: l
\end{ecode}
%
which triggers an occurs-check error in the ``cons'' case.
%
We then instructed four annotators (one of whom is an author, the other
three are teaching assistants at UCSD) to classify the answers as
correct or incorrect.

We performed an inter-rater reliability (IRR) analysis to determine the
degree to which the annotators consistently graded the exams.
%
As we had more than two annotators assigning nominal (``correct'' or
``incorrect'') ratings we used Fleiss' kappa~\cite{Fleiss1971-du} to
measure IRR.\@
%
Fleiss' kappa is measured on a scale from $1$, indicating total
agreement, to $-1$, indicating total disagreement, with $0$ indicating
random agreement.

\paragraph{Results}
%
Figure~\ref{fig:results-user-study} summarizes a single annotator's
results, which show that students given \toolname's jump-compressed
trace were consistently more likely to correctly explain and fix the
type error than those given \ocaml's error message.
%
The measured kappa values were $\kappa = 0.72$ for the explanations and
$\kappa = 0.83$ for the fixes; while there is no formal notion for what
consititutes strong agreement~\cite{Krippendorff2012-wd}, kappa values
above $0.60$ are often called ``substantial''
agreement~\cite{Landis1977-ey}.
%
\begin{figure*}[ht]
\centering
\includegraphics[width=0.49\linewidth]{user-study-reason.png}
\includegraphics[width=0.49\linewidth]{user-study-fix.png}
\caption{A classification of students' explanations and fixes for type
  errors, given either \ocaml's error % message
  or \toolname's
  jump-compressed trace. The students given \toolname's jump-compressed
  trace consistently scored better than those given \ocaml's type
  error.}
\label{fig:results-user-study}
\end{figure*}
%

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors: (1) with high coverage in a timespan amenable to
compile-time analysis, (2) with traces that have a low average
complexity of 7 jumps, and (3) that are more helpful to novice
programmers than traditional type error messages.

### 2016-06-15 paper/icfp16-cameraready/gallery2.tex
finish draft of user study

\subsection{The Advantage of Traces}\label{sec:advantage-traces}

@@

\paragraph{Example: Recursion with Bad Base Case}
%
The function @sumList@ should add up
the elements of its input list.
%
\begin{ecode}
  let rec sumList xs = match xs with
    | []    -> ==[]==
    | y::ys -> y + __sumList ys__
\end{ecode}
%
Unfortunately, in the base case, it returns @[]@
instead of @0@.
%
\sherrloc\ blames the base case, and \ocaml\
assumes the base case is correct and blames
the \emph{recursive call} on line 3:
%
\begin{verbatim}
  This expression has type
    'a list
  but an expression was expected of type
    int
\end{verbatim}
%
Both of the above are parts of the full story, which
is succinctly summarized by \toolname's trace showing
how @sumList [1; 2]@ gets stuck at @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
The trace clarifies immediately (via the third step)
that the @[]@ is the result of the recursive call
@sumList []@, and shows how it is incompatible with
the subsequent \texttt{+} operation.

@@

\paragraph{Example: Bad Helper Function that Type-Checks}
%
The function @digitsOfInt@ is supposed to return a list of
the digits of the input integer.
%
\begin{ecode}
  let append x xs =
    match xs with
    | [] -> [x]
    | _  -> x :: xs

>>>>

\subsection{Witness Utility}\label{sec:advantage-traces}

@@

\paragraph{Example: Recursion with Bad Base Case}
%
The function @sumList@ should add up
the elements of its input list.
%
\begin{ecode}
  let rec sumList xs = match xs with
    | []    -> ==[]==
    | y::ys -> y + __sumList ys__
\end{ecode}
%
Unfortunately, in the base case, it returns @[]@
instead of @0@.
%
\sherrloc\ blames the base case, and \ocaml\
assumes the base case is correct and blames
the \emph{recursive call} on line 3:
%
\begin{verbatim}
  This expression has type
    'a list
  but an expression was expected of type
    int
\end{verbatim}
%
Both of the above are parts of the full story, which
is summarized by \toolname's trace showing
how @sumList [1; 2]@ gets stuck at @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
The trace clarifies immediately (via the third step)
that the @[]@ is the result of the recursive call
@sumList []@, and shows how it is incompatible with
the subsequent \texttt{+} operation.

@@

\pagebreak
\paragraph{Example: Bad Helper Function that Type-Checks}
%
The function @digitsOfInt@ is supposed to return a list of
the digits of the input integer.
%
\begin{ecode}
  let append x xs =
    match xs with
    | [] -> [x]
    | _  -> x :: xs
